Multi-atlas Segmentation of the Whole Hippocampus and Subfields Using Multiple Automatically Generated Templates

### 2014-04-09 paper.Rnw
Other reviewer comments from R2

In this paper we describe a thorough validation of the \mb{} algorithm for the
fully automatic segmentation of the hippocampus and its subfields. First, we
address the very idea of boostrapping a template library from a limited number
of input atlases \citep{Chakravarty2013} for whole hippocampus segmentation by
conducting a multi-fold validation experiment over a range of atlas and template
library sizes, registration and label fusion methods. This type of validation is
done first on a subset of the Alzheimer's Disease Neuroimaging Initiative (ADNI)
dataset with manual segmentations  \prsnr{}-protocol, and then replicated on a
first episode psychosis patient dataset to determine the behaviour of \mb{} when
segmenting younger and differently diseased subjects. Next, we compare \mb{}
with other popular segmentation algorithms (\freesurfer{}, \fsl{}, \maper{}, and
\snt{}) on all the images available in the \adnidataset{} sample. Lastly, using
the optimal parameter settings for \mb{} found from the previous experiments, we
investigate hippocampal subfield segmentation by conducting a leave-one-out
validation using the \citet{Winterburn2013} manually segmented high-resolution
MR atlases.

@@

\mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) bootstraps the creation of a large template library given a limited
input atlas library, and then uses the template library in basic \multiatlas{}
segmentation. Images for the template library are selected from a set of input
target images, either arbitrarily or so as to reflect the neuroanatomy or
demographics of the target set as a whole (for instance, by sampling equally
from cases and controls). The template library images are automatically
labelled by each of the atlases via label propagation. Effectively, basic
\multiatlas{} segmentation is then conducted using the template library to
segment the entire set of target images (including the target images used in the
construction of the template library). Since each template library image has
multiple labels (one from each atlas), the final number of labels to be fused
for each target may be quite large (i.e. \# of atlas $\times$ \# of templates).

@@

In this experiment we explore the very idea of bootstrapping a template library
for multi-atlas-based segmentation from a small number of input atlases. To do
so, we conduct repeated cross-validations of \mb{} whilst varying the
composition and sizes of the atlas and template libraries used, as well as
varying the registration algorithm and label fusion method. The dataset used in
this experiment is images from the ADNI dataset \citep{Jack2008} along with
whole hippocampus labels manually segmented following the \prsnr-protocol
\citep{Pruessner2000}.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 4: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expsubfield{}: \expsubfielddesc{}}
\label{exp:winterburn-xval}
The previous experiment assesses \mb{} performance on whole hippocampus
segmentation. In this experiment, we evaluate \mb{} hippocampal subfield
segmentation of standard 3T T1-weighted images at $0.9mm$-isotropic voxels. We
use a modified leave-one-out cross-validation (LOOCV) design.

@@

\paragraph{Evaluation method}  
Evaluating the agreement of automated hippocampal subfield segmentations with
manual segmentations for T1 images at $0.9mm$-isotropic voxels is inherently
ill-defined since there are no manual protocols for segmentation at this
resolution. Instead, we must evaluate the reliability, or {\it precision}, with
which \mb{} produces hippocampal subfields segmentations at this resolution that
correspond in form to the segmentation protocol used by the given
high-resolution atlas library images.

By directly resampling the Winterburn atlas segmentations to $0.9mm^3$ voxels
(using standard nearest-neighbour image resampling techniques) we obtain a
subsampled version of the labels which preserve the original segmentation
protocol within the limits of error from rounding and interpolation. Therefore,
using the resampled Winterburn segmentations as definitive for the $0.9mm^3$
resolution we evaluate reliability of \mb{} segmentations using DSC overlap
scores and evaluate consistency across the range of hippocampal sizes using
Bland-Altman plots of subfield volumes.

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in order
to provide accurate and reliable segmentations with respect to manual
segmentations. Both \expadnixval{} (Section \ref{exp:adni-xval}) and
\expfepxval{} (Section \ref{exp:fep-xval}) compare \mb{} to basic-\multiatlas{}
segmentation by characterising the change in segmentation quality with varying
parameter settings (atlas and template library sizes, registration method, and
label fusion method) and differing age and neuropsychiatric populations.
Together, these experiments allow us to choose optimal \mb{} parameter settings
for use in subsequent experiments. \expadni{} (Section \ref{exp:adni})
demonstrates that across \Sexpr{length(means.complete$RID)} images from the
\adnidataset{} dataset, \mb{} performs as well as, or better, than other
established and popular methods, and has a much more conservative proportional
bias in segmentation volume.  Finally, \expsubfield{} (Section
\ref{exp:winterburn-xval}) demonstrates the reliability of \mb{} in producing
subfield segmentations which match the segmentation protocol of the input
atlases despite contrast and resolution limitations in standard T1-weighted
image volumes.  All of these experiments together demonstrate that \mb's
algorithmic performance is not dependent on a single definition of the
hippocampus but is effective with differing hippocampal definitions
\citep{Winterburn2013,Pruessner2000,Hsu2002}, across image types, and subject
populations.

The core claim the \mb{} method is based on -- that we can meaningfully bootstrap
a template library from a small set of labelled atlas images -- is validated in
the cross validation conducted in \expadnixval{} (and the replication in
\expfepxval{} and \expadnixvalsnt{}, Supplementary Materials). We find that both
increasing the number of atlases and the number of templates used improves \mb{}
segmentation over and above basic-\multiatlas{} segmentations using the same
number of atlas images. That is, by taking the extra step of generating a
template library using target images, \mb{} is able to improve the overlap
between the automatically generated segmentations and manually generated ``gold
standard'' segmentations. The magnitude of this improvement is greatest with a
small number of atlases, but even with larger atlas libraries we have found that
generating a template library reduces the variability in segmentation precision
(i.e. \mb{} more consistently produces high quality segmentations than does
basic-\multiatlas{} segmentation). These effects do not appear dependant on the
hippocampal segmentation protocol used.

@@

Among the automated segmentation methods we compared in this paper
(\freesurfer{}, \maper{}, \fsl{}), we find extremely variable
performance of all methods. With the exception of \fsl{} all methods correlate
well with the \snt{} volumes provided in the ADNI database. However, \freesurfer{}
and \fsl{} provide radically different definitions of the size of the hippocampus
in comparison to the other methods. Further, when estimating bias of these
methods relative to \snt{} hippocampal volumes we see that large hippocampi are
over estimated while small hippocampi are under estimated. By comparison, \mb{}
and \maper{} are far more conservative in volume estimation, suggesting these
methods may be better suited for estimating true-positives, especially in
neurodegenerative disease subjects featuring smaller overall hippocampi.
However, in this analysis we have only compared methods by total hippocampal
volume, and so more work is needed to understand the full extent to which these
methods differ.

@@

In conclusion, we have demonstrated the viability of leveraging a small
number of input atlases to bootstrap a large template library and thereby
improve segmentation reliability when using \multiatlas{} methods. We
demonstrated that this method works robustly over hippocampal definitions,
different disease populations, and different acquisition types. Finally, we also
demonstrate that reliable reproduction of hippocampal subfield segmentations in
standard 3T T1-weighted images is possible.

>>>>

In this paper we describe a thorough validation of the \mb{} algorithm for the
fully automatic segmentation of the hippocampus and a proof-of-concept
validation of its application to the segmentation of hippocampal subfields in
standard T1-weighted images. First, we address the very idea of generating a
template library from a limited number of input atlases \citep{Chakravarty2013}
for whole hippocampus segmentation by conducting a multi-fold validation
experiment over a range of atlas and template library sizes, registration and
label fusion methods. This type of validation is done first on a subset of the
Alzheimer's Disease Neuroimaging Initiative (ADNI) dataset with manual
segmentations  \prsnr{}-protocol, and then replicated on a first episode
psychosis patient dataset to determine the behaviour of \mb{} when segmenting
younger and differently diseased subjects. Next, we compare \mb{} with other
popular segmentation algorithms (\freesurfer{}, \fsl{}, \maper{}, and \snt{})
on all the images available in the \adnidataset{} sample. Lastly, using the
optimal parameter settings for \mb{} found from the previous experiments, we
investigate hippocampal subfield segmentation by conducting a leave-one-out
validation using the \citet{Winterburn2013} manually segmented high-resolution
MR atlases.

@@

The \mb{} ({\textbf M}ultiple {\textbf A}utomatically {\textbf Ge}nerated {\textbf
T}emplates) algorithm creates a large template library given a much smaller
sized input atlas library and then uses this template library in basic
\multiatlas{} segmentation to segment a set of input target images. The images
used in the template library are selected from the input images, either
arbitrarily or so as to reflect the neuroanatomy or demographics of the target
set as a whole (for instance, by sampling equally from cases and controls). The
template library images are automatically labelled by each of the atlases via
label propagation.  Effectively, basic \multiatlas{} segmentation is then
conducted using the template library to segment the entire set of target images
(including the target images used in the construction of the template library).
Since each template library image has multiple labels (one from each atlas),
the final number of labels to be fused for each target may be quite large (i.e.
\# of atlas $\times$ \# of templates).

@@

In this experiment we explore the very idea of generating a template library
for multi-atlas-based segmentation from a small number of input atlases. To do
so, we conduct repeated cross-validations of \mb{} whilst varying the
composition and sizes of the atlas and template libraries used, as well as
varying the registration algorithm and label fusion method. The dataset used in
this experiment is images from the ADNI dataset \citep{Jack2008} along with
whole hippocampus labels manually segmented following the \prsnr-protocol
\citep{Pruessner2000}.

@@

\FloatBarrier
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%                                                                     %%%%%
%%%%%        Experiment 4: Hippocampal Subfield Cross-Validation          %%%%%
%%%%%                                                                     %%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\expsubfield{}: \expsubfielddesc{}}
\label{exp:winterburn-xval}
The previous experiment assesses \mb{} performance on whole hippocampus
segmentation. In this experiment, we conduct a proof-of-concept evaluatation of
\mb{} hippocampal subfield segmentation of standard 3T T1-weighted images at
$0.9mm$-isotropic voxels. We use a modified leave-one-out cross-validation
(LOOCV) design.

@@

\paragraph{Evaluation method}  
Evaluating the agreement of automated hippocampal subfield segmentations with
manual segmentations for T1 images at $0.9mm$-isotropic voxels is inherently
ill-defined since there are no manual protocols for segmentation at this
resolution. Instead, we must evaluate how well the lower-resolution \mb{}
hippocampal subfield segmentations correspond in form to the segmentation
protocol used in the high-resolution images. By directly resampling the
Winterburn atlas segmentations to $0.9mm^3$ voxels (using standard
nearest-neighbour image resampling techniques) we obtain a subsampled version
of the labels which preserve the original segmentation protocol within the
limits of error from rounding and interpolation. Therefore, using the resampled
Winterburn segmentations as definitive for the $0.9mm^3$ resolution we evaluate
agreement of \mb{} segmentations using DSC overlap scores and evaluate
consistency across the range of hippocampal sizes using Bland-Altman plots of
subfield volumes.

@@

In this manuscript we have presented the implementation and validation of the
\mb{} framework -- a methodology that requires very few input atlases in order
to provide accurate and reliable segmentations with respect to manual
segmentations. Both \expadnixval{} (Section \ref{exp:adni-xval}) and
\expfepxval{} (Section \ref{exp:fep-xval}) compare \mb{} to basic-\multiatlas{}
segmentation by characterising the change in segmentation quality with varying
parameter settings (atlas and template library sizes, registration method, and
label fusion method) and differing age and neuropsychiatric populations.
Together, these experiments allow us to choose optimal \mb{} parameter settings
for use in subsequent experiments. \expadni{} (Section \ref{exp:adni})
demonstrates that across \Sexpr{length(means.complete$RID)} images from the
\adnidataset{} dataset, \mb{} performs as well as, or better, than other
established and popular methods, and has a much more conservative proportional
bias in segmentation volume.  Finally, \expsubfield{} (Section
\ref{exp:winterburn-xval}) is a proof-of-concept validation demonstrating the
reliability of \mb{} in producing subfield segmentations which match the
segmentation protocol of the input atlases despite contrast and resolution
limitations in standard T1-weighted image volumes.  All of these experiments
together demonstrate that \mb's algorithmic performance is not dependent on a
single definition of the hippocampus but is effective with differing
hippocampal definitions \citep{Winterburn2013,Pruessner2000,Hsu2002}, across
image types, and subject populations.

The core claim the \mb{} method is based on -- that a useful template library
can be generated from a small set of labelled atlas images -- is validated in
the cross-validation conducted in \expadnixval{} (and the replication in
\expfepxval{} and \expadnixvalsnt{}, Supplementary Materials).  We find that
both increasing the number of atlases and the number of templates used improves
\mb{} segmentation over and above basic-\multiatlas{} segmentations using the
same number of atlas images. That is, by taking the extra step of generating a
template library using target images, \mb{} is able to improve the overlap
between the automatically generated segmentations and manually generated ``gold
standard'' segmentations. The magnitude of this improvement is greatest with a
small number of atlases, but even with larger atlas libraries we have found
that generating a template library reduces the variability in segmentation
agreement (i.e. \mb{} more consistently produces segmentations in greater
agreement with manual segmentations than does basic-\multiatlas{} method, over
repeated randomized trials). These effects do not appear dependant on the
hippocampal segmentation protocol used.

@@

Among the automated segmentation methods we compared in this paper
(\freesurfer{}, \maper{}, \fsl{}), we find extremely variable
performance of all methods. With the exception of \fsl{} all methods correlate
well with the semi-automated \snt{} volumes provided in the ADNI database.
However, the \freesurfer{} and \fsl{} hippocampal segmentations are on average
about twice the volume of those from all other methods. Furthermore, when
estimating the bias of \freesurfer{} and \fsl{} relative to the \snt{}
hippocampal volumes we see that large hippocampi are over estimated while small
hippocampi are under estimated. By comparison, \mb{} and \maper{} are far more
conservative in volume estimation, suggesting these methods may be better
suited for estimating true-positives, especially in neurodegenerative disease
subjects featuring smaller overall hippocampi.  However, in this analysis we
have only compared methods by total hippocampal volume, and so more work is
needed to understand the full extent to which these methods differ.

@@

In conclusion, we have demonstrated the viability of leveraging a small
number of input atlases to generate a large template library and thereby
improve segmentation reliability when using \multiatlas{} methods. We
demonstrated that this method works robustly over hippocampal definitions,
different disease populations, and different acquisition types. Finally, we also
demonstrate that reliable reproduction of hippocampal subfield segmentations in
standard 3T T1-weighted images is possible.
